An Answer Set Programming Framework for Reasoning about Agents' Beliefs and Truthfulness of Statements

### 2020-06-18 main.tex
Naming of bodies of rules clarified.

A set of rules $RD$ is \emph{consistent} if, for any consistent set of fluent literals $X$, the logic program 
$\{b \leftarrow body  \mid rule(r, b, body) \in RD\} \cup \{l \leftarrow \:\mid\: l\in X\}$
is consistent.
%
\begin{definition}
A $KB = \langle O, RD \rangle$ is \emph{consistent} if $O$ and $RD$ are consistent. 
\end{definition} 
%
Our goal is to identify the agent's beliefs and to use them to determine whether  a statement $stm(p, s)$ is true at a time step $t$ given $KB$. We will develop a program $\Pi(KB)$ to answer this question. We assume a finite horizon of steps, denoted by $s(0), \ldots, s(k)$. We use $h(L,S)$ to encode that fluent literal $L$ is true at step $S$. We use $believes(B,S)$ to state that the agent believes $B$ at step $S$. Furthermore, to simplify the use of $\Pi(KB)$ with current ASP solvers, we  encode a rule of the form \eqref{rule}  using {\bf (a)} the ASP atom $rule(r, head, name)$ and
{\bf (b)} the set of ASP atoms $\{mbr(l, r) \mid l \in body_{r}\}$, where \textit{name} is the name for the rule and $mbr$ stands for `\textit{member\_of}'; similar encodings are used for  \eqref{default} and \eqref{pref}. 
This method of encoding is similar to what is used  in answer set planning and allows for the separation between  the encoding of the problem and the rules for reasoning with it.
The main challenge to be addressed lies in the interaction between observations and defaults. 
We apply the following principles:
%
\begin{itemize}
\item[]{\bf (O)}ptimistic: The most recent observation reflects the true state of the world;
\item[]{\bf (SK)}eptical: When conflicting conclusions can be drawn then do not believe in any.
\end{itemize}

@@

In this section, we shift our focus to the set $A$ of actions from the action signature $\langle F, B, A \rangle$. Following common practice in action languages \cite{GelfondL98}, we assume that each action is  associated with a set of literals, called its \emph{effects}, with each effect potentially predicated on a different set of literals, called \emph{preconditions}. A third set of literals, the \emph{executability conditions}, states when the action can be executed vs impossible. This information is encoded in statements of the form: 
$exec(a,body)$ and $causes(a,p,body)$ 
%\begin{align}
% exec(a,body) \label{exec} \\
% causes(a,p,body) \label{effect} 
%\end{align}
%
where $a \in A$ is an action, $p$ is a fluent literal, and $body$ is a set of fluent literals. 
 The first statement gives the executability conditions of action $a$, while the second states that, when $body$ holds, $p$ is one effect of $a$. We represent an \emph{action occurrence observation} (or {\em action occurrence}) by a statement of the form
 $occ(a,s)$
%\begin{equation}\label{occu}
%occ(a,s)
%\end{equation}
where $a\in A$ and $s$ is a time step. Recall that the notion of knowledge base already supports providing action occurrence observations, as  $H^{n}$ is a set of observations and action occurrences. The semantics of the above statements is encoded by suitable rules added to $\Pi(KB)$. 
 Note that the body of $exec$ and $causes$ statements is encoded by means of the $mbr$ predicate, as we did for rules and defaults in the previous section. 
We add to $\Pi(KB)$ the following rules:
 %and can be encoded   as follows. 
%
\begin{eqnarray}
obs(P, S)  & \leftarrow & s(S), occ(A, S), exec(A, M), \nonumber    \\
& & mbr(P, M).  \label{lp:exec} \\  
obs(L, S+1) & \leftarrow & s(S),     occ(A, S), causes(A, L, M), \nonumber    \\
& &  N   =\# count \{X : mbr(X, M)\},   \nonumber\\
& & \# count \{X : mbr(X, M),  \nonumber \\ 
&& \hspace*{.4in} h(X, T) \} == N.  \label{lp:effect} 
\end{eqnarray}

>>>>

A set of rules $RD$ is \emph{consistent} if, for any consistent set of fluent literals $X$, the logic program 
$\{b \leftarrow body  \mid rule(r, b, body) \in RD\} \cup \{l \leftarrow \:\mid\: l\in X\}$
is consistent.
%
\begin{definition}
A $KB = \langle O, RD \rangle$ is \emph{consistent} if $O$ and $RD$ are consistent. 
\end{definition} 
%
Our goal is to identify the agent's beliefs and to use them to determine whether  a statement $stm(p, s)$ is true at a time step $t$ given $KB$. We will develop a program $\Pi(KB)$ to answer this question. We assume a finite horizon of steps, denoted by $s(0), \ldots, s(k)$. We use $h(L,S)$ to encode that fluent literal $L$ is true at step $S$. We use $believes(B,S)$ to state that the agent believes $B$ at step $S$. Furthermore, to simplify the use of $\Pi(KB)$ with current ASP solvers, we  encode a rule of the form \eqref{rule}  using {\bf (a)} the ASP atom $rule(r, head, body\_id_{r})$ and
{\bf (b)} the set of ASP atoms $\{mbr(l, body\_id_r) \mid l \in body_{r}\}$, where $body\_id$ is an identifier for the body of the rule and $mbr$ stands for `\textit{member\_of}'; similar encodings are used for  \eqref{default} and \eqref{pref}. 
This method of encoding is similar to what is used  in answer set planning and allows for the separation between  the encoding of the problem and the rules for reasoning with it.
The main challenge to be addressed lies in the interaction between observations and defaults. 
We apply the following principles:
%
\begin{itemize}
\item[]{\bf (O)}ptimistic: The most recent observation reflects the true state of the world;
\item[]{\bf (SK)}eptical: When conflicting conclusions can be drawn then do not believe in any.
\end{itemize}

@@

In this section, we shift our focus to the set $A$ of actions from the action signature $\langle F, B, A \rangle$. Following common practice in action languages \cite{GelfondL98}, we assume that each action is  associated with a set of literals, called its \emph{effects}, with each effect potentially predicated on a different set of literals, called \emph{preconditions}. A third set of literals, the \emph{executability conditions}, states when the action can be executed vs impossible. This information is encoded in statements of the form: 
$exec(a,body)$ and $causes(a,p,body)$ 
%\begin{align}
% exec(a,body) \label{exec} \\
% causes(a,p,body) \label{effect} 
%\end{align}
%
where $a \in A$ is an action, $p$ is a fluent literal, and $body$ is a set of fluent literals. 
 The first statement gives the executability conditions of action $a$, while the second states that, when $body$ holds, $p$ is one effect of $a$. We represent an \emph{action occurrence observation} (or {\em action occurrence}) by a statement of the form
 $occ(a,s)$
%\begin{equation}\label{occu}
%occ(a,s)
%\end{equation}
where $a\in A$ and $s$ is a time step. Recall that the notion of knowledge base already supports providing action occurrence observations, as  $H^{n}$ is a set of observations and action occurrences. The semantics of the above statements is encoded by suitable rules added to $\Pi(KB)$. 
 Note that the body of $exec$ and $causes$ statements is encoded by means of the $mbr$ predicate, as we did for rules and defaults in the previous section. 
We add to $\Pi(KB)$ the following rules:
 %and can be encoded   as follows. 
%
\begin{eqnarray}
obs(P, S)  & \leftarrow & s(S), occ(A, S), exec(A, M), \nonumber    \\
& & mbr(P, M).  \label{lp:exec} \\  
obs(L, S+1) & \leftarrow & s(S),     occ(A, S), causes(A, L, M), \nonumber    \\
& &  N   =\# count \{X : mbr(X, M)\},   \nonumber\\
& & \# count \{X : mbr(X, M),  \nonumber \\ 
&& \hspace*{.4in} h(X, S) \} == N.  \label{lp:effect} 
\end{eqnarray}
